Program Transformations for miniKanren

### 2020-07-05 papers/aplas/text/evaluation.tex
Evaluation intro section rewritten

\section{Evaluation}
In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
For some programs we have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.
We measured the execution time of each specialized program for some queries and compared it with the execution time of the original program.

We used the following programs to test the specializers on.
\begin{itemize}
  \item A program to compute a concatenation of three lists.
  \item A program to compute both the length of the list and its maximal element.
  \item Two implementations of an evaluator of logic formulas.
  \item Two implementations of a typechecker for a small expression language.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

The last two relations are described in~\cite{lozov2019relational} thus we will only briefly describe them in this paper.

We focused on these particular examples because they are examples of relational interpreters which are natural and observable.
In this study we only measured the execution time for sample queries.
The measured time was averaged over multiple runs for every query.
The queries were run on a laptop running Ubuntu 18.04 with quad core Intel Core i5 2.30GHz CPU and 8 GB of RAM.

\todo{intro sentence.}
Row Original contains the execution time of the original version of the program before any transformations.
Row Ecce corresponds to the result of the transformation with Ecce conjunctive partial deduction system.
Row Non-CPD --- translation by our implementation of nonconjunctive partial deduction.
Etalon --- the etalon implementation of the same program.
CPD --- the earlier implementation of the conjunctive partial deduction for \mk{}.
Brancing --- the earlier implemenctation of CPD for \mk{} with the branching heuristic.

@@

The better implementation of this relation (see fig.~\ref{doubleApp:etalon}) does not traverse the first list twice.
This etalon implementation is generated with the conjunctive partial deduction.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.6\textwidth}
    \begin{lstlisting}[label={doubleApp:etalon}, caption={Etalon implementation of concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res = conde [
    (xs === [] /\ append$^o$ ys zs res);
    fresh (h t r) (
      xs === h % t /\
      res === h % r /\
      doubleAppend$^o$ t ys zs r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

There is no significant difference in the execution time in the backward direction between different specialized versions~\ref{tbl:doubleApp}.
However the execution time in the backward direction differs: the program generated by Ecce shows the same speedup as the etalon program, while non-conjunctive partial deduction slows the program down.
\todo{WHY?}

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & forward & backward \\
  \hline\hline
  Original         & 0.0040s & 0.0195s \\ \hline
  Ecce             & 0.0031s & 0.0201s \\ \hline
  Non-CPD          & 0.0055s & 0.0202s \\ \hline
  Etalon           & 0.0031s & 0.0203s \\
  \end{tabular}
  \caption{Evaluation results for doubleAppendo}
  \label{tbl:doubleApp}
\end{table}

@@

The etalon implementation is shown in fig.~\ref{etalon:maxandlength}.
It uses a recursive relation to compute the length and maximum element simultaneously while traversing the list once.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
\begin{lstlisting}[label={etalon:maxandlength}, caption={Etalon implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l = conde [
    (xs === [] /\ m === n /\ l === zero);
    (fresh (h t l$_1$)
       (xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
Note, that the second disjunct in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Thus among others we compared two etalon implentations: one includes these disjuncts, while the other removes them (see table~\ref{tbl:maxlen})

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 2.286ms  \\ \hline
  Etalon           & 4.623ms  \\ \hline
  Etalon removed   & 3.133ms  \\ \hline
  Ecce             & 1.701ms  \\ \hline
  Non-CPD          & 2.578ms
  \end{tabular}

@@

Surprising enough, the execution time of the etalon program is worse than of the original.
Ecce succeeds at improving the program performance, while non-conjunctive partial deduction worsens it a little.
\todo{why}

@@

All examples of \mk{} relations in this paper are written in \oc{}\footnote{\oc{}: statically typed \mk{} embedding in \ocaml{}. The repository of the project: \url{https://github.com/JetBrains-Research/OCanren}} syntax.
We specialize the \lstinline{eval$^o$} relation to synthesize formulas which evaluate to \lstinline{^true}\footnote{An arrow lifts ordinary values to the logic domain}.
To do so, we run the specializer for the goal with the last argument fixed to \lstinline{^true}, while the first two arguments remain free variables.
Depending on the way the \lstinline{eval$^o$} is implemented, different specializers generate significantly different residual programs.

@@

One possible implementation of the evaluator in the syntax of \oc{} is presented in listing~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean operations.

@@

Note, that the calls to boolean relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} are placed last within each conjunction.
This poses a challenge to the CPD-based specializers.
Conjunctive partial deduction unfolds relation calls from left to right, so when specializing this relation for running backwards (i.e. considering the goal \lstinline{eval$^o$ subst fm ^true}), it fails to propagate the direction data onto recursive calls of \lstinline{eval$^o$}.
Knowing that \lstinline{res} is \lstinline{^true}, we can conclude that in the call \lstinline{and$^o$ v w res} variables \lstinline{v} and \lstinline{w} have to be \lstinline{^true} as well.
There are three possible options for these variables in the call \lstinline{or$^o$ v w res} and one for the call \lstinline{not$^o$}.
These variables are used in recursive calls of \lstinline{eval$^o$} and thus restrict the result of driving them.
CPD fails to recognize this, and thus unfolds recursive calls of \lstinline{eval$^o$} applied to fresh variables.
It leads to over-unfolding, big residual programs and less than optimal performance.

@@

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & last    & plain  \\
  \hline\hline
  Original         & 0.695s  & 1.426s \\
  \hline
  CPD              & 1.721s  & 0.168s \\
  \hline
  Ecce             & 0.205s  & 0.264s \\
  \hline
  Non-CPD          & 0.132s  & 0.137s \\
  \hline
  Branching        & 0.468s  & 0.179s \\
  \end{tabular}

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk{} program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by CPD (0.352 seconds) while the program generated by adding branching heuristic also fails to terminate in 30 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 1.947 seconds.
While driving this program, the non-conjunctive partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

@@

All evaluation results are presented in the table~\ref{tbl:unify}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk{} relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c||c}
                   & \multicolumn{3}{c||}{unify} & isPath \\
  \hline
                   & first  & second & third     &        \\
  \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$ (oum?) & $>30s$ (oum?) & $23.743s$ \\
  \hline
  CPD              & $0.081*10^{-4}s$ & $0.017s$ & $0.343s$  & $2.999s$  \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
  \hline
  Ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
  \hline
  Branching        & $0.205*10^{-4}s$ & $0.101s$ & $>30s$ (oum?) & N/A \\
  \hline
  \end{tabular}

@@

@@

%% \begin{align*}
%%   &type \ term = \\
%%   &| \ BConst \ of \ Bool \\
%%   &| \ IConst \ of \ Int \\
%%   &| \ Var \ of \ Int \\
%%   &| \ term + term \\
%%   &| \ term * term \\
%%   &| \ term = term \\
%%   &| \ term < term \\
%%   &| \ \underline{let} \ term \ \underline{in} \ term \\
%%   &| \ \underline{if} \ term \ \underline{then} \ term \ \underline{else} \ term
%% \end{align*}

@@

We compared two possible implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational}.
The second program is hand-written in \oc{}.
Each implementation has been transformed with non-conjunctive partial deducer and by Ecce.

We measured the time needed to generate 100 terms of type integer (averaged over 10 iterations).
Ecce significantly worsens the execution time for both implementations.
Non-conjunctive partial deduction improves the first implementation a little bit, while worsening the performance of the second implementation 3 times.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\ \hline\hline
  Original    & 0.464s           & 0.057s           \\ \hline
  Non-CPD     & 0.448s           & 0.154s           \\ \hline
  Ecce        & 1.518s           & 2.543s           \\
  \end{tabular}

>>>>

\section{Evaluation}
In our study we compared the new non-conjunctive partial deduction with the \ecce partial deduction system.
\ecce is designed for \pro programming language and cannot be directly applied to \mk.
To be able to compare our approach with \ecce, we converted each input program first to the pure subset of \pro, then transformed it with \ecce, and then we converted the result back to the pure subset of \pro.
The coversion to \pro is a simple syntactic conversion.
In conversion from \pro to \mk, a conjunction is generated by each horn clause in which unifications are placed before any relation call.
This mimics in \mk the execution order of \pro program within a conjunction.

The set of benchmarks consist of both examples from the literature on partial deduction and programs of our own creation.
The first two programs are well-known in conjunctive partial deduction literature~\cite{de1999conjunctive}.
The second two are introduced in the paper on relational interpreters~\cite{lozov2019relational}.
The last two are examples of relational interpreters created by us which are observable, but non-trivial.
In the last two examples we explore how different implementations of the same relation affects the quality of transformations.

The \lstinline{doubleAppend$^o$} program concatenates three lists of numbers by computing a conjunction of two calls of a relation which concatenates two lists.
This example is prominent in CPD literature, since it demonstrates \emph{deforestation} --- a transformation which gets rid of intermidiate datastructures.
This transformation cannot be done with the earlier partial deduction approach, since the relation uses a variable shared between two relation calls within a conjunction which requires the transformation to treat conjunctions as a whole.
We explore how the fact that our approach sometimes splits conjunctions ignoring the variable sharing affects the quality of transformation.

The \lstinline{maxLength$^o$} program computes the maximum element of the list as well as its length.
In this program an input list is shared between the call of the relation to compute the maximum element and the relation to compute the length of the list.
This sharing leads to the list being traversed twice during execution.
This example is used to demonstrate \emph{tupling}: the transformation in which multiple traversals of the same data structure are replaced with a single traversal which computes all the necessary results simultaneously.
CPD performes tupling more often than our approach, and we explore how it affects the performance of the transformed program.

The \lstinline{isPath} and \lstinline{unify} relations demonstrate the relation interpretation approach in which a relation is generated from a functional program.
Partial deduction is capable to improve the execution time of the generated relations in the given direction.
These two relations are described in~\cite{lozov2019relational} thus we will only briefly describe them in this paper.
Here the aim is to study if our approach can help to achieve comparable performance improvement as \ecce.

The \lstinline{eval$^o$} relation implements an evaluator of a subset of propositional formulas.
We consider four different implementations of this relation to explore how the way program is implemented can affect the quality of transformation.
Depending on the implementation, \ecce generates programs of varying performance, while the execution time of the program generated by our approach is similar.

The \lstinline{typecheck$^o$} relation implements a typechecker for a tiny expression language.
We consider two different implementations of this relation: one written by hand and the other generated from the functional program.
We demonstrate how much these implementations differ in terms of performance before and after transformations.

In this study we only measured the execution time for the sample queries, averaging them over multiple runs.
All examples of \mk relations in this paper are written in \oc\footnote{\oc: statically typed \mk embedding in \ocaml. The repository of the project: \url{https://github.com/JetBrains-Research/OCanren}}.
The queries were run on a laptop running Ubuntu 18.04 with quad core Intel Core i5 2.30GHz CPU and 8 GB of RAM.

The tables and graphs use the following denotations.
\emph{Original} represents the execution time of a program before any transformations were applied; \emph{ECCE} --- of the program transformed with \ecce; \emph{Non-CPD} --- of the program transformed by our approach.
The first two examples also contain \emph{Ideal} row which contain the execution time of the ideal implementations of the relations presented in the paper~\cite{de1999conjunctive}.

@@

The better implementation of this relation (see fig.~\ref{doubleApp:ideal}) does not traverse the first list twice.
This ideal implementation is generated with the conjunctive partial deduction.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.6\textwidth}
    \begin{lstlisting}[label={doubleApp:ideal}, caption={Ideal implementation of concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res = conde [
    (xs === [] /\ append$^o$ ys zs res);
    fresh (h t r) (
      xs === h % t /\
      res === h % r /\
      doubleAppend$^o$ t ys zs r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

There is no significant difference in the execution time in the backward direction between different specialized versions~\ref{tbl:doubleApp}.
However the execution time in the backward direction differs: the program generated by \ecce shows the same speedup as the ideal program, while non-conjunctive partial deduction slows the program down.
\todo{WHY?}

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & forward & backward \\
  \hline\hline
  Original         & 0.0040s & 0.0195s \\ \hline
  \ecce             & 0.0031s & 0.0201s \\ \hline
  Non-CPD          & 0.0055s & 0.0202s \\ \hline
  Ideal           & 0.0031s & 0.0203s \\
  \end{tabular}
  \caption{Evaluation results for doubleAppendo}
  \label{tbl:doubleApp}
\end{table}

@@

The ideal implementation is shown in fig.~\ref{ideal:maxandlength}.
It uses a recursive relation to compute the length and maximum element simultaneously while traversing the list once.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
\begin{lstlisting}[label={ideal:maxandlength}, caption={Ideal implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l = conde [
    (xs === [] /\ m === n /\ l === zero);
    (fresh (h t l$_1$)
       (xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
Note, that the second disjunct in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Thus among others we compared two ideal implentations: one includes these disjuncts, while the other removes them (see table~\ref{tbl:maxlen})

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 2.286ms  \\ \hline
  Ideal           & 4.623ms  \\ \hline
  Ideal removed   & 3.133ms  \\ \hline
  \ecce             & 1.701ms  \\ \hline
  Non-CPD          & 2.578ms
  \end{tabular}

@@

Surprising enough, the execution time of the ideal program is worse than of the original.
\ecce succeeds at improving the program performance, while non-conjunctive partial deduction worsens it a little.
\todo{why}

@@

We specialize the \lstinline{eval$^o$} relation to synthesize formulas which evaluate to \lstinline{^true}\footnote{An arrow lifts ordinary values to the logic domain}.
To do so, we run the specializer for the goal with the last argument fixed to \lstinline{^true}, while the first two arguments remain free variables.
Depending on the way the \lstinline{eval$^o$} is implemented, different specializers generate significantly different residual programs.

@@

One possible implementation of the evaluator in the syntax of \oc is presented in listing~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean operations.

@@

Note, that the calls to boolean relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} are placed last within each conjunction.
This poses a challenge to the CPD-based specializers.
Conjunctive partial deduction unfolds relation calls from left to right, so when specializing this relation for running backwards (i.e. considering the goal \lstinline{eval$^o$ subst fm ^true}), it fails to propagate the direction data onto recursive calls of \lstinline{eval$^o$}.
Knowing that \lstinline{res} is \lstinline{^true}, we can conclude that in the call \lstinline{and$^o$ v w res} variables \lstinline{v} and \lstinline{w} have to be \lstinline{^true} as well.
There are three possible options for these variables in the call \lstinline{or$^o$ v w res} and one for the call \lstinline{not$^o$}.
These variables are used in recursive calls of \lstinline{eval$^o$} and thus restrict the result of driving them.
CPD fails to recognize this, and thus unfolds recursive calls of \lstinline{eval$^o$} applied to fresh variables.
It leads to over-unfolding, big residual programs and poor performance.

@@

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & last    & plain  \\
  \hline\hline
  Original         & 0.695s  & 1.426s \\
  \hline
  \ecce             & 0.205s  & 0.264s \\
  \hline
  Non-CPD          & 0.132s  & 0.137s \\
  \end{tabular}

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by CPD (0.352 seconds) while the program generated by adding branching heuristic also fails to terminate in 30 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 1.947 seconds.
While driving this program, the non-conjunctive partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

@@

All evaluation results are presented in the table~\ref{tbl:unify}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c}
    \multirow{ 2}{*}{Terms} &
    f(X, a) & f(a \% b \% nil, c \% d \% nil, L) & f(X, X, g(Z, t))  \\
    \cline{2-4} &
    f(a, X) & f(X \% XS, YS, X \% ZS) & f(g(p, L), Y, Y)  \\
    \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$   \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  \\
  \hline
  \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$ \\
  \hline
  \end{tabular}

@@

% \begin{table}
%   \centering
%   \begin{tabular}{c||c|c|c||c}
%                    & \multicolumn{3}{c||}{unify} & isPath \\
%   \hline
%                    & first  & second & third     &        \\
%   \hline\hline
%   Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$  & $23.743s$ \\
%   \hline
%   Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
%   \hline
%   \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
%   \hline
%   \end{tabular}

%   \caption{Evaluation results of unification and path search}
%   \label{tbl:unify}
% \end{table}

@@

@@

We compared two possible implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational}.
The second program is hand-written in \oc.
Each implementation has been transformed with non-conjunctive partial deducer and by \ecce.

We measured the time needed to generate 100 terms of type integer (averaged over 10 iterations).
\ecce significantly worsens the execution time for both implementations.
Non-conjunctive partial deduction improves the first implementation a little bit, while worsening the performance of the second implementation 3 times.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\ \hline\hline
  Original    & 0.464s           & 0.057s           \\ \hline
  Non-CPD     & 0.448s           & 0.154s           \\ \hline
  \ecce        & 1.518s           & 2.543s           \\
  \end{tabular}

### 2020-07-05 papers/aplas/text/intro.tex
Evaluation intro section rewritten

The core feature of the family of relational programming languages \mk{}\footnote{\mk{} language web site: \url{http://minikanren.org}} is their ability to run a program in different directions.
Having specified a relation for adding two numbers, one can also compute the subtraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running \emph{backwards} a relational interpreter for some language.
In general, it is possible to create a solver for a recognizer by translating it into \mk{} and running in the appropriate direction~\cite{lozov2019relational}.

The search employed in \mk{} is complete which means that every answer will be found, although it may take a long time.
The promise of \mk{} falls short when speaking of performance.
The running time of a program in \mk{} is highly unpredictable and varies greatly for different directions.
What is even worse, it depends on the order of the relation calls within a program.
One order can be good for one direction, but slow down the computation drastically in the other direction.

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving the performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of a relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} showed that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk{} programs.
Unfortunately, it may also not affect the running time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro{} have been studied before~\cite{leuschel2002logic}.
The ideas described there are aimed at left-to-right evaluation strategy of \pro{}.
Since the search in \mk{} is complete, it is safe to reorder some relation calls within the goal for better performance.
While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk{}.
We compare it to the existing specialization algorithms on several programs and discuss why some \mk{} programs run slower after specialization.

>>>>

The core feature of the family of relational programming languages \mk\footnote{\mk language web site: \url{http://minikanren.org}} is their ability to run a program in different directions.
Having specified a relation for adding two numbers, one can also compute the subtraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running \emph{backwards} a relational interpreter for some language.
In general, it is possible to create a solver for a recognizer by translating it into \mk and running in the appropriate direction~\cite{lozov2019relational}.

The search employed in \mk is complete which means that every answer will be found, although it may take a long time.
The promise of \mk falls short when speaking of performance.
The running time of a program in \mk is highly unpredictable and varies greatly for different directions.
What is even worse, it depends on the order of the relation calls within a program.
One order can be good for one direction, but slow down the computation drastically in the other direction.

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving the performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of a relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} showed that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk programs.
Unfortunately, it may also not affect the running time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro have been studied before~\cite{leuschel2002logic}.
The ideas described there are aimed at left-to-right evaluation strategy of \pro.
Since the search in \mk is complete, it is safe to reorder some relation calls within the goal for better performance.
While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk.
We compare it to the existing specialization algorithms on several programs and discuss why some \mk programs run slower after specialization.

### 2020-07-05 papers/aplas/text/nonconj.tex
Move APLAS draft to another folder
Evaluation intro section rewritten

In this section, we describe a novel approach to relational programs specialization.
This approach draws inspiration from both conjunctive partial deduction and supercompilation.
The aim was to create a specialization algorithm which is simpler than conjunctive partial deduction and uses properties of \mk{} to improve the performance of the input programs.

@@

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into the scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Other disjuncts take the form of possibly empty conjunction of relation calls accompanied with a substitution computed from unifications.
Any \mk{} term can be trivially transformed into the described form.
In Fig.~\ref{fig:ncpd-pseudo} function \code{normalize} is assumed to perform term normalization.
The code is omitted for brevity.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.85\textwidth]{algo_pseudo-crop.pdf}
  \caption{\db{Non-conjunctive Partial Deduction Pseudo Code}}
  \label{fig:ncpd-pseudo}
\end{figure}

@@

When the heuristic selects a call to unfold (line 15), a process tree is constructed for the selected call \emph{in isolation} (line 16).
The leaves of the computed tree are examined.
If all leaves are either computed substitutions or are instances of some relations accompanied with non-empty substitutions, then the leaves are collected and each of them replaces the considered call in the root conjunction (lines 19--20).
If the selected call does not suit the criteria, the results of its unfolding are not propagated onto other relation calls within the conjunction, instead, the next suitable call is selected (line 22).
According to the denotational semantics of \mk{} it is safe to compute individual conjuncts in any order, thus it is ok to drive any call and then propagate its results onto the other calls.

>>>>

In this section, we describe a novel approach to relational programs specialization.
This approach draws inspiration from both conjunctive partial deduction and supercompilation.
The aim was to create a specialization algorithm which is simpler than conjunctive partial deduction and uses properties of \mk to improve the performance of the input programs.

@@

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into the scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Other disjuncts take the form of possibly empty conjunction of relation calls accompanied with a substitution computed from unifications.
Any \mk term can be trivially transformed into the described form.
In Fig.~\ref{fig:ncpd-pseudo} function \code{normalize} is assumed to perform term normalization.
The code is omitted for brevity.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.85\textwidth]{algo_pseudo-crop.pdf}
  \caption{Non-conjunctive Partial Deduction Pseudo Code}
  \label{fig:ncpd-pseudo}
\end{figure}

@@

When the heuristic selects a call to unfold (line 15), a process tree is constructed for the selected call \emph{in isolation} (line 16).
The leaves of the computed tree are examined.
If all leaves are either computed substitutions or are instances of some relations accompanied with non-empty substitutions, then the leaves are collected and each of them replaces the considered call in the root conjunction (lines 19--20).
If the selected call does not suit the criteria, the results of its unfolding are not propagated onto other relation calls within the conjunction, instead, the next suitable call is selected (line 22).
According to the denotational semantics of \mk it is safe to compute individual conjuncts in any order, thus it is ok to drive any call and then propagate its results onto the other calls.

### 2020-07-05 papers/aplas/text/related.tex
Evaluation intro section rewritten

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for left-to-right evaluation strategy of \pro{}.
CPD constructs a tree which models goal evaluation and is similar to SLD-NF tree, then a residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the local control determines the shape of the residual programs, while the global control ensures that every relation which can be called in the residual program is indeed defined.
The leaves of local control trees become nodes of the global control tree.
CPD analyses these nodes at the global level and runs local control for all those which are new.

At the local level, CPD examines a conjunction of atoms by considering each atom one-by-one from left to right.
An atom is \emph{unfolded} if it is deemed safe, i.e. a whistle based on homeomorphic does not signal for the atom.
When an atom is unfolded, a clause whose head can be unified with the atom is found, and a new node is added into the tree where the atom in the conjunction is replaced with the body of that clause.
If there is more than one suitable head, then several branches are added into the tree which corresponds to the disjunction in the residualized program.
An adaptation of CPD for the \mk{} programming language is described in~\cite{lozov2019relational}.

The most well-behaved strategy of local control in CPD for \pro{} is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if only one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within conjunction, it is most likely to be unfolded and introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk{}.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro{} because it mimics the way programs in \pro{} execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk{} program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.

>>>>

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for left-to-right evaluation strategy of \pro.
CPD constructs a tree which models goal evaluation and is similar to SLD-NF tree, then a residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the local control determines the shape of the residual programs, while the global control ensures that every relation which can be called in the residual program is indeed defined.
The leaves of local control trees become nodes of the global control tree.
CPD analyses these nodes at the global level and runs local control for all those which are new.

At the local level, CPD examines a conjunction of atoms by considering each atom one-by-one from left to right.
An atom is \emph{unfolded} if it is deemed safe, i.e. a whistle based on homeomorphic does not signal for the atom.
When an atom is unfolded, a clause whose head can be unified with the atom is found, and a new node is added into the tree where the atom in the conjunction is replaced with the body of that clause.
If there is more than one suitable head, then several branches are added into the tree which corresponds to the disjunction in the residualized program.
An adaptation of CPD for the \mk programming language is described in~\cite{lozov2019relational}.

The most well-behaved strategy of local control in CPD for \pro is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if only one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within conjunction, it is most likely to be unfolded and introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro because it mimics the way programs in \pro execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.
